Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
minor fixes, American English, quote marks, etc.

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom (UK). The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point 
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, quoting Kersten Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{\url{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

@@

Despite these differences, and despite the fact that software often has taken
the role of a proper, albeit less tangible, scientific research instrument,
neither the software nor its creators are commonly credited as part of a
scientific publication. The group acknowledged the need for more recognition for
the creators of such software instruments, and indicated a number of possible
pathways. First and foremost, domain scientists must be made aware of the
important role of software, and include the developers as co-authors of papers.
A second approach is to fully embrace an open badging infrastructure (such as
Mozilla's Open Badges), where a {\em badge\/} is a free, transferrable,
evidence-based indicator of an accomplishment, skill, quality, or interest. A
third approach is for the scientific community to support the increasing
momentum of peer-reviewed journals specialized in the open source/open access
publication of scientific research software, such as Computer Physics
Communication, F1000 Research, Journal of Open Research Software, and SoftwareX.

@@

Discovery of relevant software is still fundamental issue: we need standard vocabularies and metadata, better tools, and approaches that are sustainable. Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of "the community"? Some communities have had significant initiatives to improve software discovery, e.g. the NIH Software Discovery Index\footnote{NIH Software Discovery Index: \url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of journals which had made software more discoverable: ACM Transactions on Software offer a reproducibility review; the Journal of Biostatistics has an opt-in to provide code and a certification mark if it can be run; the Journal of Open Research Software requires software to be deposited in suitable repositories and referenced.

@@

Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website. Using code repositories like GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a licence or by archiving (with a DOI) in Figshare\footnote{FigShare: \url{http://figshare.com/}} or Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}. Docker\footnote{Docker: \url{htp://www.docker.com/}} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.

@@

\subsubsection{Second breakout discussion: Categorisation of software journals}

This group aimed to come up with a categorisation of journals which published software papers. Starting from the list of journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers. These were the Journal of Open Research Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}}, PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}}, Journal of Statistical Software\footnote{\url{http://www.jstatsoft.org/instructions}}. Methods in Ecology and Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}}, Transactions of Mathematical Software\footnote{\url{http://toms.acm.org/Authors.html}}, GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS Computational Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.

@@

\item Software Policies
\begin{itemize}
	\item Software must have a license
	\begin{itemize}
		\item Software must have an open source license
	\end{itemize}
	\item Availability
	\begin{itemize}
		\item Software must be openly available and accessible
	\end{itemize}
	\item Deposit policies
	\begin{itemize}
		\item Software should be in a public repository
		\begin{itemize}
			\item Of particular stature / with preservation plan
		\end{itemize}
		\item Software should have a permanent identifier
		\item Software deposit doesn?t count as a prior publication
	\end{itemize}
	\item Runnability and dependencies
	\begin{itemize}
		\item Provide documentation to understand how to run
		\item Provide sample data
		\item Provide dependency information
	\end{itemize}
\end{itemize}
\end{itemize}

@@

Journals should provide guidelines about what to consider when reviewing a software submission. A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}. Journals might also learn from organisations such as the Apache Foundation as to what is good practice for software submissions. Guidance is needed on what constitutes an incremental improvement that is significant enough to qualify for publication, otherwise this assessment can be very subjective.

@@

\begin{table*}[t]
\centering
\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
  \begin{scriptsize}
  \begin{tabular}{l|l|r|r}
 \hline
    Author  &   Tweet  & Retweets &  Favorites
\\ \hline 
% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    & 
%
% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
%\\     &  management plan used by EPSRC software grants    &    &  
% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    & 
%
 Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
\\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    & 
\\   &   we provide a software evaluation tool:   &    & 
\\   &   \url{http://www.software.ac.uk/online-sustainability-evaluation}  &    & 
%
\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
\\ & UK researchers write research software w$/$out any formal training &    & 
%
\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
%
\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    & 
%
\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
\\ &   [include] how many faces are up on our community page''   &    & 
%
\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
\\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    & 
% 
%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
% 
\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & & 
%
\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
\\ &   \url{http://www.slideshare.net/kaythaney/}     & & 
\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
%
\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & & 
%
%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 & 
%
\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
\\ & but want SWORD/DSpace/Fedora/ePrints see:  & & 
\\ & \url{http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/}  & & 
%
\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
\\ & Bioinformatics Manifesto: \url{https://github.com/pjotrp/bioinformatics}   & 
%
\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3 
\\ & but because influential people are averse to change...''  C. Sunstein     & & 
% 
\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
\\   &  Research Software \url{http://arxiv.org/abs/1407.5981}  & & 
%
\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
%
\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
\\ &   $+$ sustainability. Tune in! Live stream: \url{http://ustre.am/17ddh} & & 
%
\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe2/agenda/}   &  &
\\ & URL for live stream of keynotes \& lightning talks: \url{http://ustre.am/17ddh}   &  &
\\ \hline
    \end{tabular}
    \end{scriptsize}
\end{table*}

>>>>

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the
United Kingdom (UK). The SSI was founded to support the UK's research software
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, quoting Kerstin Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{\url{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

@@

Despite these differences, and despite the fact that software often has taken
the role of a proper, albeit less tangible, scientific research instrument,
neither the software nor its creators are commonly credited as part of a
scientific publication. The group acknowledged the need for more recognition for
the creators of such software instruments, and indicated a number of possible
pathways. First and foremost, domain scientists must be made aware of the
important role of software, and include the developers as co-authors of papers.
A second approach is to fully embrace an open badging infrastructure (such as
Mozilla's Open Badges), where a {\em badge} is a free, transferrable,
evidence-based indicator of an accomplishment, skill, quality, or interest. A
third approach is for the scientific community to support the increasing
momentum of peer-reviewed journals specialized in the open source/open access
publication of scientific research software, such as Computer Physics
Communication, F1000 Research, Journal of Open Research Software, and SoftwareX.

@@

Discovery of relevant software is still fundamental issue: we need standard vocabularies and metadata, better tools, and approaches that are sustainable. Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of ``the community''? Some communities have had significant initiatives to improve software discovery, e.g. the NIH Software Discovery Index\footnote{NIH Software Discovery Index: \url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of journals which had made software more discoverable: ACM Transactions on Software offer a reproducibility review; the Journal of Biostatistics has an opt-in to provide code and a certification mark if it can be run; the Journal of Open Research Software requires software to be deposited in suitable repositories and referenced.

@@

Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website. Using code repositories like GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a license or by archiving (with a DOI) in Figshare\footnote{FigShare: \url{http://figshare.com/}} or Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}. Docker\footnote{Docker: \url{htp://www.docker.com/}} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.

@@

\subsubsection{Second breakout discussion: Categorization of software journals}

This group aimed to come up with a categorization of journals which published software papers. Starting from the list of journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers. These were the Journal of Open Research Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}}, PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}}, Journal of Statistical Software\footnote{\url{http://www.jstatsoft.org/instructions}}. Methods in Ecology and Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}}, Transactions of Mathematical Software\footnote{\url{http://toms.acm.org/Authors.html}}, GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS Computational Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.

@@

\item Software Policies
\begin{itemize}
	\item Software must have a license
	\begin{itemize}
		\item Software must have an open source license
	\end{itemize}
	\item Availability
	\begin{itemize}
		\item Software must be openly available and accessible
	\end{itemize}
	\item Deposit policies
	\begin{itemize}
		\item Software should be in a public repository
		\begin{itemize}
			\item Of particular stature / with preservation plan
		\end{itemize}
		\item Software should have a permanent identifier
		\item Software deposit doesn't count as a prior publication
	\end{itemize}
	\item Runnability and dependencies
	\begin{itemize}
		\item Provide documentation to understand how to run
		\item Provide sample data
		\item Provide dependency information
	\end{itemize}
\end{itemize}
\end{itemize}

@@

Journals should provide guidelines about what to consider when reviewing a software submission. A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}. Journals might also learn from organizations such as the Apache Foundation as to what is good practice for software submissions. Guidance is needed on what constitutes an incremental improvement that is significant enough to qualify for publication, otherwise this assessment can be very subjective.

@@

\begin{table*}[t]
\centering
\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
  \begin{scriptsize}
  \begin{tabular}{l|l|r|r}
 \hline
    Author  &   Tweet  & Retweets &  Favorites
\\ \hline
% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    &
%
% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
%\\     &  management plan used by EPSRC software grants    &    &
% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    &
%
 Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
\\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    &
\\   &   we provide a software evaluation tool:   &    &
\\   &   \url{http://www.software.ac.uk/online-sustainability-evaluation}  &    &
%
\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
\\ & UK researchers write research software w$/$out any formal training &    &
%
\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
%
\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    &
%
\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
\\ &   [include] how many faces are up on our community page''   &    &
%
\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
\\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    &
%
%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
%
\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & &
%
\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & &
\\ &   \url{http://www.slideshare.net/kaythaney/}     & &
\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & &
%
\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & &
%
%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 &
%
\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
\\ & but want SWORD/DSpace/Fedora/ePrints see:  & &
\\ & \url{http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/}  & &
%
\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
\\ & Bioinformatics Manifesto: \url{https://github.com/pjotrp/bioinformatics}   &
%
\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3
\\ & but because influential people are averse to change...''  C. Sunstein     & &
%
\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
\\   &  Research Software \url{http://arxiv.org/abs/1407.5981}  & &
%
\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
%
\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
\\ &   $+$ sustainability. Tune in! Live stream: \url{http://ustre.am/17ddh} & &
%
\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe2/agenda/}   &  &
\\ & URL for live stream of keynotes \& lightning talks: \url{http://ustre.am/17ddh}   &  &
\\ \hline
    \end{tabular}
    \end{scriptsize}
\end{table*}
